Mining of Keystroke and Mouse Dynamics to Increase the Engagement of Students with Programming Assignments

### 2018-04-20 paper.tex
Extends the state of the art

Although the use of KD is found in several research works
as a biometric measure, its use for identifying affective states is rare in
comparison. Epp, Lippold \& Mandryk \cite{epp2011identifying} effectively used KD in conjunction
with decision-tree classifiers for the identification of 15 affective states.
Although their work was based on fixed text, their technique to extract a feature
vector was an inspiration for the proposed method in this work. As for free-text
KD, Bixler y D\'Mello \cite{bixler2013detecting} present a method for the identification of boredom
and engagement based on several classification models. A similar situation exists in the case of Mouse Dynamics (MD), which is used mainly for authentication; however, Salmeron-Majadas, Santos
and Boticario \cite{salmeron2014exploring} use both MD and KD to predict four affective states using
five different classification algorithms. Bakhtiyari y Husain \cite{bakhtiyari2014fuzzy} discuss a
method based on fuzzy models for the recognition of emotions through KD, MD and
touch-screen interactions. For a broad review of emotion recognition methods
based on KD and MD, the work by Kolakowska \cite{kolakowska2013review} is recommended.

>>>>

Although the use of KD is found in several research works
as a biometric measure, its use for identifying affective states is rare in
comparison. Epp, Lippold \& Mandryk \cite{epp2011identifying} effectively used KD in conjunction
with decision-tree classifiers for the identification of 15 affective states.
Although their work was based on fixed text, their technique to extract a feature
vector was an inspiration for the proposed method in this work. As for free-text
KD, Bixler y D\'Mello \cite{bixler2013detecting} present a method for the identification of boredom
and engagement based on several classification models. A similar situation exists in the case of Mouse Dynamics (MD), which is used mainly for authentication; however, Salmeron-Majadas, Santos
and Boticario \cite{salmeron2014exploring} use both MD and KD to predict four affective states using
five different classification algorithms. Bakhtiyari y Husain \cite{bakhtiyari2014fuzzy} discuss a
method based on fuzzy models for the recognition of emotions through KD, MD and
touch-screen interactions. Lim \cite{lim2017detecting} use several
sensor sets and methods for detecting stress levels, including neural
networks; the detected stress level is then fed into the inference
system that decides on adaptation. However, this work focuses only on
stress level, disregarding the more general affective state. Although
it is no longer current, the state of the art was described quite
comprehensively in Kolakowska's paper \cite{kolakowska2013review}.

More recent papers, like the one by Carneiro and Novais \cite{CARNEIRO2017171} takes a
more comprehensive approach, using also a non-intrusive system based
on keystroke analysis; several affective states are measured: mental
fatigue, stress and emotional valence. However, it is not focused
particularly in programming, which has certain characteristics, but
uses music to create a certain mood on the subjects. The KD
methodology, however, is applicable to a wide range of
situations. Also Kolakowska has recently published a paper
\cite{Kolakowska2018} assessing the usefulness of KD not only in
emotion recognition, but also in user authentication. Whatever the
method, it is indeed proved that KD can be used for detecting a range
of user mood and attitudes \cite{wrobel2018applicability}. In this
paper we are more concerned with the data collection and the selection
of a particular method for detecting the mood with the limited amount
of data obtained via KD.
